---
title: "Day 1 - File 4"
author: "Melinda Higgins"
date: "7/28/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# set up knitr options
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# load libraries needed for this document
library(dplyr)
library(Hmisc)
library(readr)

# load the abalone dataset
abalone <- readr::read_csv("abalone.csv")
# add sex as a factor type variable
abalone$sex_factor <- factor(abalone$sex,
                             levels = c("M","F","I"),
                             labels = c("Male",
                                        "Female",
                                        "Immature"))

```

Let's make some other cool tables. For the exercises below, you will be exploring these other packages:

* **tableone**
* **table1**
* **arsenal**

## Try out a cool package with the **table1** package

This package makes HTML tables.

Here is the syntax code for creating a table of summary statistics for the variables of diameter, height and length by sex for the abalone dataset.

```{r}
library(table1)
table1(~diameter + height + length | sex,
       data=abalone)
```

## Descriptive Data - a useful blog article

Learn more at [http://thatdatatho.com/2018/08/20/easily-create-descriptive-summary-statistic-tables-r-studio/](http://thatdatatho.com/2018/08/20/easily-create-descriptive-summary-statistic-tables-r-studio/).

### Let's look at the `table1` package - for HTML tables

The package vignette is [https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html](https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html)

We can add some labels and make another table.

```{r}
table1::label(abalone$shuckedWeight) <- "Shucked Weight"
table1::label(abalone$visceraWeight) <- "Viscera Weight"
table1::label(abalone$shellWeight) <- "Shell Weight"
 
table1::table1(~shuckedWeight + 
                 visceraWeight + 
                 shellWeight | sex, 
               data = abalone)
```

### Next try the `tableone` package

See package vignette [https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html](https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html)

```{r}
library(tableone)
CreateTableOne(data = abalone)
```

### Also try the `arsenal` package

See more at [https://eheinzen.github.io/arsenal/articles/tableby.html](https://eheinzen.github.io/arsenal/articles/tableby.html)

**NOTE**: We have to add `results="asis"` to the code chunk options for this to look pretty in the output.

Let's make a table of the descriptive statistics for Abalone dimensions by sex. Look at height, diameter and length.

```{r results="asis"}
library(arsenal)
table_one <- tableby(sex ~ height + diameter + length, data = abalone)
summary(table_one, title = "Abalone Data")
```

## Your turn

Create a table of the abalone weights by sex. Put in variables wholeWeight, shuckedWeight, visceraWeight and shellWeight by sex.

Change the title to "Descriptive statistics of Abalone weights by sex".

```{r results="asis"}
library(arsenal)
table_one <- 
  tableby(sex ~ wholeWeight + shuckedWeight +
            visceraWeight + shellWeight, 
          data = abalone)
summary(table_one, 
        title = "Descriptive statistics of Abalone weights by sex")
```

## Try different output formats

Knit this document "to HTML", then "to PDF", and then "to Word" (DOC) formats and see what happens.

