---
title: "Day 1 - File 3"
author: "Melinda Higgins"
date: "7/28/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# set up knitr options
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# load libraries needed for this document
library(dplyr)
library(Hmisc)
library(readr)

# load the abalone dataset
abalone <- readr::read_csv("abalone.csv")
# add sex as a factor type variable
abalone$sex_factor <- factor(abalone$sex,
                             levels = c("M","F","I"),
                             labels = c("Male",
                                        "Female",
                                        "Immature"))

```

## Get specific statistics on the abalones

There are many functions for getting specific statistics such as: `mean()`, `sd()`, `median()`, `min()`, `max()`, and `quantile()`. Each of these functions assumes only 1 variable (vector) is being input to the function. So, let's use the `pull()` function from the **dplyr** package.

Get the mean diameter of all of the abalones.

```{r}
abalone %>%
  pull(diameter) %>%
  mean()
```

Let's get the mean and standard deviation for the abalones diameter. We'll use the `summarise()` function from **dplyr**.

```{r}
abalone %>%
  summarise(mean = mean(diameter), 
            sd = sd(diameter))
```

Let's get the median and IQR. With the updated version of **dplyr** 1.0.0 we can get multiple values at once.

HINT: Type `sessionInfo()` to get the current listing of all package loaded into memory at this point in your analysis.

```{r}
sessionInfo()
```

Get the median and IQR of the abalone diameters.

```{r}
abalone %>%
  summarise(median = median(diameter),
            iqr = quantile(diameter, c(0.25, 0.75)))
```

## Other options for multiple stats for multiple variables

Let's use the `summarise_all()` function with the `list()` option for multiple statistical functions.

```{r}
abalone %>%
  select(length, height, diameter) %>%
  summarise_all(list(mean, sd))
```

## Grouping

Another useful function is the `group_by()` function from **dplyr**. Let's get these same stats for these variables again, but this time for each abalone sex. 

So, we need to add the `group_by()` function to create a "grouped" data class that these **dplyr** functions use to group the output.

This is similar to a CLASS statement in SAS.

```{r}
abalone %>%
  group_by(sex) %>%
  select(sex, length, height, diameter) %>%
  summarise_all(list(mean, sd))
```

## Your turn

Get the median and iqr for the abalone wholeWeight by sex - use the `summarise()` function here, not the `summarise_all()` function.

```{r}
# insert your code here
abalone %>%
  group_by(sex) %>%
  summarise(median = median(diameter),
            iqr = quantile(diameter,(c(0.25, 0.75))))

```

