---
title: "Day 1 - File 2 - cleaner output"
author: "Melinda Higgins"
date: "7/28/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # echo=TRUE shows R code, FALSE hide code
knitr::opts_chunk$set(error = TRUE) # error=TRUE document compile even with errors
knitr::opts_chunk$set(message = FALSE) # do not show messages
knitr::opts_chunk$set(warning = FALSE) # do not show warnings
```

```{r loaddata}
# load the abalone dataset
library(readr)
abalone <- readr::read_csv("abalone.csv")
```

## Get some summary statistics

Let's get some summary statistics using the `summary()` function. The summary function will give us the mean, sd, median, min and max for numerical data. _All that is missing is the interquartile range._

```{r getsummary}
library(dplyr)
# get summary stats of diameter
abalone %>%
  select(diameter) %>%
  summary()
```

However, the `summary()` function is not very useful for character data.

```{r}
# get summary stats of sex - frequencies
abalone %>%
  select(sex) %>%
  summary()
```

We could create a "factor" type variable for `sex`.

```{r}
abalone$sex_factor <- factor(abalone$sex,
                             levels = c("M","F","I"),
                             labels = c("Male",
                                        "Female",
                                        "Immature"))
abalone %>%
  select(sex_factor) %>%
  summary()
```

Let's try some other packages and functions. The `describe()` function is really useful from the **Hmisc** package.

```{r}
library(Hmisc)
# get descriptive stats for wholeWeight
abalone %>%
  select(wholeWeight) %>%
  Hmisc::describe()

# Hmisc::describe() also works for character variables
abalone %>%
  select(sex) %>%
  Hmisc::describe()
```

Why am I typing the `packagename::functionname()`?

Some packages have functions with the SAME NAME! It is really helpful to type the name of the package followed by 2 colons :: and then the function name to avoid confusion.

For example, let's try the `describe()` from the **psych** package.

```{r}
library(psych)
abalone %>%
  select(wholeWeight) %>%
  psych::describe()
```

Oops we get an error. What happened??

It turns out it is often important to know what each function is producing.

Let's save and review the results from the `select()` function above.

```{r}
# save this output in an object named aww_select
aww_select <- abalone %>%
  select(wholeWeight)
# what is the class of this object?
class(aww_select)
```

This is a "tibble data frame". The **psych** package `describe()` function doesn't know what to do with this kind of data object.

There is another way to pull out 1 variable. Use the `pull()` function.

Let's see what kind of object that creates.

```{r}
# save this output in an object named aww_pull
aww_pull <- abalone %>%
  pull(wholeWeight)
# what is the class of this object?
class(aww_pull)
```

This is now a numeric vector.

Let's try our `describe()` function from the **psych** package again with this approach.

```{r}
abalone %>%
  pull(wholeWeight) %>%
  psych::describe()
```

We can also use the `as.data.frame()` function to convert the output to a regular data.frame object and then call `psych::describe()`.

There are a number of `as.xxx()` and `is.xxx()` functions which are helpful for sorting out "class-type mismatch" errors that often occur in R. 

```{r}
abalone %>%
  select(wholeWeight) %>%
  as.data.frame() %>%
  psych::describe()
```

It is worth noting that the `psych::describe()` function only works on numeric data. It will not work on character variables nor factor type variables.

Type `help(describe, package="psych")` in the console.

```{r}
abalone %>%
  pull(sex) %>%
  psych::describe()

abalone %>%
  pull(sex_factor) %>%
  psych::describe()
```

## This file won't knit??

Notice that this file will not knit. This is because some of our code above has errors. In production we probably do not want to create a report with errors in it. So, by default **Rmarkdown** won't compile or "knit" a document that has errors.

We can override this by adding an **knitr** option setting `error = TRUE` and the document should knit.

Add this to the 1st code chunk above and try knitting this document again.

```
knitr::opts_chunk$set(error = TRUE)
```

## Make it a little cleaner

We can add more **knitr** options to remove the warnings and other messages that are cluttering up our document.

Here are 2 more options to add to the 1st code chunk above. Add these and re-knit the document to see the difference.

```
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

And what happens if you set `echo=FALSE`?

Learn more about knitr options at [https://yihui.org/knitr/options/](https://yihui.org/knitr/options/).