---
title: "Day 2 - File 02 - Data Import and Export"
author: "Melinda Higgins"
date: "7/29/2020"
output:
  html_document: default
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# set up knitr options
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# load libraries needed for this document
library(dplyr)
library(Hmisc)
library(readr)
```

## Loading .RData files

When you load in the native binary .RData formatted file, the objects stored in that datafile all get loaded into your R environment. When we saved this before we only saved the `abaloneMod` dataset - only 1 object.

```{r}
load("abaloneMod.RData")
```

But you can save multiple objects into one R binary datafile.

As a quick example - create 3 simple objects, a, b, c. Save all of them in an .RData file.

```{r}
a <- 4
b <- c(3,2,1)
c <- list(a,b)
a
b
c
ls()

```

Save these

```{r}
save(a,b,c, file = "abc.RData")
```

Remove these 3 objects from the environment

```{r}
rm(a,b,c)
```

Load them back in

```{r}
load("abc.RData")
ls()
```

## Other ways to import data into R

Look at the File/Import Dataset options.

Let's try the code for importing the original abalone dataset from the .SAV formatted file from SPSS.

```{r}
library(haven)
abalone <- read_sav("abalone.sav")
```

