---
title: "Day 2 - File 04 - Fit and Compare Models"
author: "Melinda Higgins"
date: "7/29/2020"
output:
  html_document: default
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# set up knitr options
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# load libraries needed for this document
library(dplyr)
library(Hmisc)
library(readr)

# load the abalone modified dataset
load("abaloneMod.RData")
```

## Let's look at some more plots

Let's look at some more plots of variables we think may be related to predicting the age of the abalones.

Let's look at age with length and whole weight.

For this let's try to **GGally** package which makes some cool plots especially matrix scatterplots. It is an extension package for **ggplot2**.

Learn more about **GGally** package at [https://cran.r-project.org/web/packages/GGally/index.html](https://cran.r-project.org/web/packages/GGally/index.html).

Also check out the "R Graph Gallery" for ideas, see [https://www.r-graph-gallery.com/199-correlation-matrix-with-ggally.html](https://www.r-graph-gallery.com/199-correlation-matrix-with-ggally.html).

```{r}
library(GGally)
abaloneMod %>% 
  select(age, length, wholeWeight) %>%
  GGally::ggpairs()
```

Color the points by sex - add this aesthetic to the `ggpairs()` function.

```{r}
abaloneMod %>% 
  select(sex, age, length, wholeWeight) %>%
  GGally::ggpairs(aes(color=sex))
```

## Fit linear models

Since I'd like to compare models, I need to work with a dataset that has no missing data. We can create a dataset with no missing values across all variables as follows.

```{r}
abaloneMod_complete <- abaloneMod %>%
  filter(complete.cases(abaloneMod))
```

Fit 3 models and save the results and look at the summary of each model.

Model 1 age by length.

```{r}
lm1 <- lm(age ~ length, data = abaloneMod_complete)
lm1
slm1 <- summary(lm1)
slm1

slm1[["coefficients"]]
```

Model 2 age by wholeWeight

```{r}
lm2 <- lm(age ~ wholeWeight, data = abaloneMod_complete)
slm2 <- summary(lm2)
slm2
```

Model 3 age by length and wholeWeight

```{r}
lm3 <- lm(age ~ length + wholeWeight,
          data=abaloneMod_complete)
slm3 <- summary(lm3)
slm3
```

Compare the full model `lm3` to the reduced models `lm1` or `lm2`.

```{r}
anova(lm1, lm3) # model with length to full model with both

anova(lm2, lm3) # model with WW to full model with both
```

## Make tables of the coefficients of each model

```{r}
library(knitr)
knitr::kable(slm1$coefficients,
             caption = "Model 1 Age by Length")
knitr::kable(slm2$coefficients,
             caption = "Model 2 Age by Whole Weight")
knitr::kable(slm3$coefficients,
             caption = "Model 3 Age by Length and Whole Weight")
```

## Another way to compare models

Use the **stargazer** package to compare models

The default output type is "latex" for making PDF documents. You will need to change this option depending on if you are knitting to HTML or to PDF. This will NOT work for WORD files.

The code chunk option must be set to `results='asis'`.

```{r results='asis'}
library(stargazer)

# set for HTML output
stargazer(lm1, lm2, lm3, type="html")

# uncomment this if knitting to PDF
# stargazer(lm1, lm2, lm3, type="latex")

# comment all of this out if knitting to WORD
```



