---
title: "Day 2 - File 01 - Data Cleaning"
author: "Melinda Higgins"
date: "7/29/2020"
output:
  html_document: default
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# set up knitr options
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# load libraries needed for this document
library(dplyr)
library(Hmisc)
library(readr)

# load the abalone dataset
abalone <- readr::read_csv("abalone.csv")
# add sex as a factor type variable
abalone$sex_factor <- factor(abalone$sex,
                             levels = c("M","F","I"),
                             labels = c("Male",
                                        "Female",
                                        "Immature"))

```

## Data errors/issues in Abalone dataset

As we saw yesterday in our Day 1 exercises, there are a few typos and errors in the Abalone dataset. Here are the key issues in this dataset:

1. There are heights = 0 which should not be possible. Perhaps this is a numerical precision issue in how the data were recorded, but without further information, 0's should not be possible. Let's set these values to `NA` (not applicable) or "missing" or "unknown".

2. As we saw yesterday there was at least 1 abalone whose `shuckedWeight` was larger than its `wholeWeight` which isn't possible. Again let's set this abalone's shucked weight and whole weight to unknown or to `NA`.

3. We also saw there were a few abalones whose length was not their longest measurement. So, again let's set these values to missing `NA`.

To make modification to the dataset, we will use the `mutate()` function from the **dplyr** package.

We will also create a new dataset called abaloneMod which has our modifications applied to this dataset. This is better than writing over the original data.

Look at case(s) where height == 0. Use the `filter()` function from **dplyr**.

NOTE: You do need to use a double equals for the logical operator of "is equal to" `==`.

```{r}
abalone %>%
  filter(height == 0)
```

We can print these "case IDs" to the screen to help us go back and check these data.

```{r}
abalone %>%
  filter(height == 0) %>%
  select(X1)
```

Using this information, create a new modified dataset `abaloneMod` that sets height equal to 0 to missing.

Note: protected words in R

TRUE
FALSE
NA
NaN

```{r}
# create abaloneMod
# set height == 0 to NA
abaloneMod <- abalone %>%
  mutate(height = ifelse(height == 0, 
                         yes = NA,     # NA is a protected word
                         no = height))

abaloneMod %>%
  filter(height == NA) %>%
  head()

abaloneMod %>%
  filter(is.na(height))
```

Let's check the stats for `height` with the summary function before and after the missing data was added for implausible values.

```{r}
# before creating the NAs
abalone %>%
  pull(height) %>%
  summary()

# after creating the NAs for height=0
abaloneMod %>%
  pull(height) %>%
  summary()
```

Notice that the output for `summary()` now shows 2 NA's - indicating we have 2 missing values for the variable `height` in the abalone dataset.

## Missing Data issues

Many R functions will throw errors if/when you have missing data. Most have options you can set to help adjust for the missing data so the function will execute and run correctly.

For example, let's see how the `mean()` function runs with missing data.

```{r}
# after creating the NAs for height=0
abaloneMod %>%
  pull(height) %>%
  mean()
```

This code runs, but the result is `NA`. Since there are missing values `NA`s for the `height` variable, the result is `NA`.

Let's add some options on how to handle the missing data. For the `mean()` function, we can add the `na.rm=TRUE` option to tell the `mean()` function to first remove the missing `NA`s and then run then calculate the mean.

"na's" and remove them = na.rm()


```{r}
# add na.rm=TRUE option
abaloneMod %>%
  pull(height) %>%
  mean(na.rm=TRUE) # remove NAs first, then compute the mean
```

Also try with `summarise()` function from **dplyr** to add additional information.

For this example below we've added 3 summary statistics.

* `n = n()` which is from the **dplyr** package to give the total count of rows of data, even with missing data
* we added `nmiss = sum(is.na(height))` which 1st runs the `is.na()` function which returns 0 for not missing and 1 for missing and then the `sum()` function counts up the number of missing values for the variable `height`
* and finally we add the `mean()` of `height` excluding the missing values. 

So, the mean height reported here is out of 4175 abalone heights with 2 abalones missing height measurements.

```{r}
# add na.rm=TRUE option
abaloneMod %>%
  select(height) %>%
  summarise(n_ht = n(),
            nmiss_ht = sum(is.na(height)),
            mean_ht = mean(height, na.rm=TRUE))
```

## More modification of the Abalone dataset

Let's add missing values for the other issues with this dataset. Set the weights to `NA` where shucked weight is > whole weight.

```{r}
# create abaloneMod
# If shuckedWeight is > wholeWeight set both values = NA
abaloneMod <- abaloneMod %>%
  mutate(shuckedWeight = 
           ifelse(shuckedWeight > wholeWeight, NA,
                  shuckedWeight),
         wholeWeight = 
           ifelse(shuckedWeight > wholeWeight, NA,
                  wholeWeight))
```

## Your turn

Get the means of `shuckedWeight` and `wholeWeight` from `abaloneMod` now that we've set these values to missing where the shucked weight was > the whole weight for that abalone.

HINT: Start with the example above:

```
abaloneMod %>%
  select(height) %>%
  summarise(n_ht = n(),
            nmiss_ht = sum(is.na(height)),
            mean_ht = mean(height, na.rm=TRUE))
```

```{r}
# put the answer here

abaloneMod %>%
  select(shuckedWeight, wholeWeight) %>%
  summarise(n_sw = n(),
            nmiss_sw = sum(is.na(shuckedWeight)),
            mean_sw = mean(shuckedWeight, na.rm=TRUE),
            n_ww = n(),
            nmiss_ww = sum(is.na(wholeWeight)),
            mean_ww = mean(wholeWeight, na.rm=TRUE))

```

## One more issue to clean up in the dataset

Let's see if/where length was perhaps shorter < than height or diameter. We can quickly view these rows (cases) using the `filter()` function from the **dplyr** package.

```{r}
# rows where length is shorter than height
abaloneMod %>%
  filter(length < height)

# rows where length is shorter than diameter
abaloneMod %>%
  filter(length < diameter)

# rows where EITHER length was shorter than height or diameter
# the | operator is used for "or"
abaloneMod %>%
  filter((length < height) | (length < diameter))

# learn more - type in help("|") to see the list of other logical operators such as ! & and more
```

Using this knowledge, let's set these dimension values to missing where either length is < height OR < diameter.

```{r}
# If (length < height) | (length < diameter) set all values = NA
abaloneMod <- abaloneMod %>%
  mutate(length = 
           ifelse((length < height) | (length < diameter), 
                  NA,
                  length),
         height = 
           ifelse((length < height) | (length < diameter), 
                  NA,
                  height),
         diameter = 
           ifelse((length < height) | (length < diameter), 
                  NA,
                  diameter))
```

And get means and n and number missing for these 3 variables adjusting for the new missing `NA` values.

```{r}
abaloneMod %>%
  select(length, height, diameter) %>%
  summarise(n_lt = n(),
            nmiss_lt = sum(is.na(length)),
            mean_lt = mean(length, na.rm=TRUE),
            nmiss_ht = sum(is.na(height)),
            mean_ht = mean(height, na.rm=TRUE),
            nmiss_dm = sum(is.na(diameter)),
            mean_dm = mean(diameter, na.rm=TRUE))
```

Now that we have missing data, let's see what a summary table of these 3 dimension variables looks like using the **arsenal** package to make a summary table.

Notice that the number of missing values is now displayed.

**Remember** Use the `abaloneMod` updated dataset.

```{r results='asis'}
library(arsenal)
tab_dimension <- 
  tableby(~length + height + diameter,
          data=abaloneMod)
summary(tab_dimension)
```

Make this table again and show the summary statistics by `sex`. Now you can see the more of the values were set to missing for the "I" infants.

```{r results='asis'}
tab_dimension <- 
  tableby(sex ~ length + height + diameter,
          data=abaloneMod)
summary(tab_dimension)
```

## One more modification - calculate age

Remember in the data notes that abalone age is equal to the number of rings + 1.5. So, let's go ahead and add `age` as a variable to our dataset computed from the number of rings + 1.5. Add this variable to our modified dataset, `abaloneMod`.

We'll again use the `mutate()` function from **dplyr**.

```{r}
abaloneMod <- abaloneMod %>%
  mutate(age = rings + 1.5)
```

Get the summary stats for `rings` and `age` by `sex` - again use the arsenal package.

```{r results='asis'}
tab_agerings <- 
  tableby(sex ~ age + rings,
          data=abaloneMod)
summary(tab_agerings)
```

## Save the update data file

To save the `abaloneMod` dataset for use in another R markdown file or other R scripts or projects, we can save it a couple of ways.

Save it as an R data file "abaloneMod.RData".

```{r}
save(abaloneMod,
     file = "abaloneMod.RData")
```

Save it as a CSV output using the `write_csv()` function from **readr** package.

```{r}
library(readr)
readr::write_csv(abaloneMod, "abaloneMod.csv")
```

