---
title: "Day 2 - File 03 - Statistical Tests"
author: "Melinda Higgins"
date: "7/29/2020"
output:
  html_document: default
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# set up knitr options
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# load libraries needed for this document
library(dplyr)
library(Hmisc)
library(readr)

# load the abalone modified dataset
load("abaloneMod.RData")
```

## Run a t-test

The function to perform a t-test is `t.test()`.

Let's look at running a t-test to compare the number of rings between males and females ignoring the infants.

Let's first create a subset of the data for just the adults.

```{r}
# select cases where sex does not equal to "I" for infants
abaloneAdults <- abaloneMod %>%
  filter(sex != "I")
```

Run t-test using the formula syntax of `rings ~ sex`.

NOTE: `var.equal=FALSE` is the default for the `t.test()` function - so by default R runs the unpooled t-test. You can change this setting to TRUE if you wish.

```{r}
t.test(rings ~ sex, 
       var.equal = TRUE,
       data=abaloneAdults)
```

This gives us the basic output for the pooled t-test, but there is a lot more we can get. So, let's save this output as an object and see what is there and how to pull it out.

```{r}
ttp <- t.test(rings ~ sex, 
              var.equal = TRUE,
              data=abaloneAdults)
ttp
```

This `ttp` object is a list of 10 other objects. Let's see what is there and how to pull out the values we might want.

```{r}
ttp$statistic    # t-test statistic
ttp$parameter    # degrees of freedom for t-test
ttp$p.value      # p-value for the t-test
ttp$conf.int     # 95% confidence interval for the mean diff
ttp$estimate     # the 2 group means for F and M
ttp$estimate[1]  # the mean of the F females
ttp$estimate[2]  # the mean of the M males
```

We could use these to help customize our output further if we wish to do so. For example, we could write the following sentence.

The pooled t-test for the difference between the average number of rings for females `r ttp$estimate[1]` and number of rings for males `r ttp$estimate[2]` was significant with a p-value of `r ttp$p.value`.

We could clean this sentence up a little bit by adding the `round()` function or the `format()` function to limit the number of decimal places shown. Let's set the number of decimal places to 4 when we round the numbers and set significant digits to 2 for the p-value.

The pooled t-test for the difference between the average number of rings for females `r round(ttp$estimate[1], digits=4)` and number of rings for males `r round(ttp$estimate[2], digits=4)` was significant with a p-value of `r format(ttp$p.value, digits=2)`.

## Your turn

Run a t-test for the abalone heights between males and females in the `abaloneAdults` dataset and display your results how ever you would like to.

```{r}
# insert your code here

```


## Let's create a median split for rings

Save the median number of rings.

```{r}
medrings <- abaloneAdults %>%
  pull(rings) %>%
  median()
```

Create a new variable for low rings and high rings. This will create a TRUE/FALSE logical variable called `highrings` for abalones with rings > 10 (the median number of rings).

```{r}
abaloneAdults <- abaloneAdults %>%
  mutate(highrings = rings > medrings)
```

Look at a table summary for `highrings`.

```{r}
abaloneAdults %>%
  pull(highrings) %>%
  table()
```

We can also look at this with `sex`.

```{r}
abaloneAdults %>%
  select(sex, highrings) %>%
  table()
```

We can make a better "Cross Table" using the gmodels package.

And we can add some options and also get a chi-square test result.

```{r}
library(gmodels)
gmodels::CrossTable(abaloneAdults$sex, 
                    abaloneAdults$highrings,
                    prop.r=FALSE,     # no row %
                    prop.c=TRUE,      # yes col %
                    prop.t=FALSE,     # no total %
                    prop.chisq=FALSE, # no chisq proportion
                    chisq=TRUE)       # yes run chisq test
```

We can also run a simple `chisq.test()` of this data and save the results like we did for the t-test.

```{r}
cs <- chisq.test(abaloneAdults$sex,
                 abaloneAdults$highrings)
```

Look at some of the values in this saved object `cs`.

```{r}
cs$statistic
cs$parameter
cs$p.value
```

Again we could write a sentence - There is a significant association between abalone sex and having the number of rings be greater than 10, p-value = `r cs$p.value`.

## Your turn

Run a chi-square test to compare abalone sex for abalones whose shell weights are < their viscera weights. I'll create the new variable for you.

```{r}
abaloneAdults <- abaloneAdults %>%
  mutate(shell_ltv = shellWeights < visceraWeight)
```

Run the chi-square test between `abaloneAdults$sex` and      `abaloneAdults$shell_ltv`.

```{r}
# insert your code here


```

## Save the modifications for `abaloneAdults`

```{r}
save(abaloneAdults, "abaloneAdults.RData")
```

